---
title: "Homework Assignment #2"
subtitle: "Exploring patterns of environmental justice"
author: "Haylee Oyler"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Part 1: Legacy of redlining in current environmental (in)justice

### Load packages
```{r}
#| code-fold: true
#| code-summary: "Reveal code"

library(sf)
library(here)
library(terra)
library(tmap)
library(tidyverse)
library(viridisLite)
library(RColorBrewer)
library(kableExtra)
library(testthat)
library(patchwork)
```

### Import data
```{r}
#| results: hide
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"), quiet = TRUE)
birds <- read_sf(here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
holc <- read_sf(here("data", "mapping-inequality", "mapping-inequality-los-angeles.json"))
```

### Filter data to desired locations
```{r}
# Filter to California
ca <- ejscreen %>% 
  filter(STATE_NAME == "California" & ! ID %in% c("060379903000","060379902000", "060599901000")) %>% 
  st_make_valid()

# Filter to LA county
la <- ejscreen %>% 
  filter(CNTY_NAME == "Los Angeles County" & ! ID %in% c("060379903000","060379902000", "060599901000"))

```

### Check CRS of the datasets we're interested in mapping
```{r}
#| warning: true
#| code-fold: true
#| code-summary: "Reveal code"

if(st_crs(la) == st_crs(holc)) {
  if(st_crs(la) == st_crs(birds)) {
    print("Coordinate reference systems match")
  }} else {
      warning("Coordinate reference systems do not match")
}
```

```{r}
# Transform CRS
holc <- st_transform(holc, crs = st_crs(la))
birds <- st_transform(birds, crs = st_crs(la))
ca <- st_transform(ca, crs = st_crs(la))
```

```{r}
#| warning: true
#| code-fold: true
#| code-summary: "Reveal code"

if(st_crs(la) == st_crs(holc)) {
  if(st_crs(la) == st_crs(birds)) {
    print("Coordinate reference systems match")
  }} else {
      warning("Coordinate reference systems do not match")
}
```

## Map of historical redlined neighborhoods in LA County
```{r}
#| code-fold: true
#| code-summary: "Reveal code"
#| label: holc-map

tm_shape(ca, bbox = holc) +
  tm_fill(col = "#f1edea") +
  tm_borders(alpha = 0.4,
             col = "grey30") +
tm_shape(la, bbox = holc) +
    tm_fill(col = "#f3ddc7") +
  tm_borders() +
tm_shape(holc) +
  tm_fill("grade",
          palette = c("#91f366", "#3472db","#f0f061", "#f54242"),
          title = "HOLC Grade",
          labels = c("A-Best", 
                     "B-Still Desirable", 
                     "C-Definitely Declining", 
                     "D-Hazardous", 
                     "Missing")) +
  tm_borders() +
tm_scale_bar(position = c("right", "top")) + # can give a vector of values 0-1
tm_compass(position = c("right", "top"),
           size = 1) +
tm_layout(main.title = "LA County historical redlined neighborhoods ",
          bg.color = "#99c2dd",
          legend.position = c(0, 0.17))
```

## Table summarizing the percent of current census block groups within each HOLC grade
```{r}
#| code-fold: true
#| code-summary: "Reveal code"

census_group <- st_join(la, holc, join = st_intersects, left = FALSE) %>% 
  group_by(grade) %>% 
  summarise(percent = n()/nrow(la)*100) %>% 
  select(grade, percent) %>% 
  st_drop_geometry()

kable(census_group, 
      digits = 2, 
      format = "simple", 
      col.names = c("Grade", "Percent of census block groups"), 
      caption = "Percent of census block groups within HOLC grades", 
      align = "l")
```
Here, I join holc data into LA county data using st_intersects and an inner join. This is because I am only interested in the redlined areas that intersect with LA county, not those outside it. 

### Join HOLC and LA county data to retrieve census block groups
```{r}
la_holc <- st_join(la, holc, join = st_intersects, left=TRUE)
```
Here, I used `st_join()` to join the redlining data into the LA county EJScreen data. I joined by `st_intersects` and I set `left=TRUE`. This allows me to see all of the redlined districts that intersect with LA county. We keep left as `TRUE` because there are some census block groups (CBGs) that have multiple HOLC assignments to them and we do not want to lose those observations.  

```{r}
#| warning: true
#| echo: false

if(nrow(la_holc) == nrow(la)){
  print("join matches original data dimensions")
} else {
  warning("join does not match orginal data dimensions")
  print(paste("la has", nrow(la), "rows"))
  print(paste("la_holc has", nrow(la_holc), "rows"))
} 
```
The rows are different because a census block group can have multiple HOLC grade assignments, so the CBGs repeat more than once in the new joined data frame. 

```{r}
# Unit tests to confirm our intuition
expect_true(length(la_holc) > length(la))
expect_equal(length(unique(la_holc$ID)), length(unique(la$ID)))
```
This confirms our intuition that the length of the joined data frame is greater than the length of the original data frame. and the length of unique CBGs in the joined data frame is the same as the length of unique CBGs in the original data frame. 

## Table summarizing adverse life outcomes by  HOLC grade
```{r}
#| code-fold: true
#| code-summary: "Reveal code"

conditions <-la_holc %>% 
  group_by(grade) %>% 
  summarise(percent_low_income = mean(LOWINCPCT, na.rm = TRUE)*100, 
            percentile_pm_25 = mean(P_PM25, na.rm = TRUE),
            percentile_low_life_expectancy = mean(P_LIFEEXPPCT, na.rm = TRUE)) %>% 
  st_drop_geometry()

kable(conditions, 
      digits = 2, 
      col.names = c("HOLC Grade", "Percent % Low Income", "Percent Percentile PM 2.5", "Percent Percentile Low Life Expectancy"), 
      format = "simple", 
      caption = "Current conditions within each HOLC redlining grade")
  
```

## Visualize the results of adverse life outcomes by HOLC grade
Here, I chose to use boxplots of the entire data frame rather than just the means because I believe it is a better visualization that is more informative of the actual distribution across grades. You can still see the median value and the IQR so you can look at trends across the grades. 
```{r}
#| code-fold: true
#| code-summary: "Reveal code"

ggplot(la_holc, aes(x = grade, y = LOWINCPCT*100)) +
  geom_boxplot(aes(fill = grade)) +
  stat_summary(fun.y=mean, 
               geom="point", 
               shape=20, 
               size=6, 
               aes(fill="mean")) +
  scale_fill_manual(values=c("#91f366", "#3472db","#f0f061", "#f54242", "grey35")) +
  labs(title = "Percentage low income by HOLC grade",
       x = "HOLC Grade",
       y = "Percentage low income (%)") +
  theme_minimal()

ggplot(la_holc, aes(x = grade, y = P_PM25)) +
  geom_boxplot(aes(fill = grade)) +
  stat_summary(fun.y=mean, 
               geom="point", 
               shape=20, 
               size=6, 
               aes(fill="mean")) +
  scale_fill_manual(values=c("#91f366", "#3472db","#f0f061", "#f54242", "grey35")) +
  labs(title = "Percentile PM 2.5 by HOLC grade",
       x = "HOLC Grade",
       y = "Percentile PM 2.5 ") +
  theme_minimal()

ggplot(la_holc, aes(x = grade, y = P_LIFEEXPPCT)) +
  geom_boxplot(aes(fill = grade)) +
  stat_summary(fun.y=mean, 
               geom="point", 
               shape=20, 
               size=6, 
               aes(fill="mean")) +
  scale_fill_manual(values=c("#91f366", "#3472db","#f0f061", "#f54242", "grey35")) +
  labs(title = "Percentile low life expectancy by HOLC grade",
       x = "HOLC Grade",
       y = "Percentile low life expectancy") +
  theme_minimal()
```
These box plots show that, on average, the adverse life and health outcomes of low income, particulate matter exposure, and low life expectancy increase as the HOLC grade decreases(gets worse). This matches with what we expect given the history of redlining and understanding how these grades were assigned as well as how the assignment continued to perpetuate and widen inequity. 

# Part 2: Legacy of redlining in biodiversity observations

### Join the bird observation data and the HOLC data
```{r}
holc_bird <- st_join(holc, birds, join = st_intersects, left=TRUE)
```
We join birds into HOLC data via a left join intersect because we want only the bird observations that are present inside the HOLC CBGs

```{r}
#| warning: true
#| echo: false

if(nrow(holc_bird) == nrow(holc)){
  print("join matches original data dimensions")
} else {
  warning("join does not match orginal data dimensions")
  print(paste("holc has", nrow(holc), "rows"))
  print(paste("holc_bird has", nrow(holc_bird), "rows"))
} 

```
The rows do not match here because we are paring down our bird observations to only those within HOLC CBGs. 

```{r}
# Adding unit tests 
expect_true(length(holc_bird) > length(holc))
expect_equal(length(unique(holc_bird$area_id)), length(unique(holc$area_id)))
```
This confirms our intuition that the joined data frame has more rows than the original HOLC data because there are multiple bird observations per HOLC area. 

## Table summarizing bird obserations by HOLC grade
```{r}
#| code-fold: true
#| code-summary: "Reveal code"

bird_final <-  holc_bird %>% 
  filter(year == 2022) %>% 
  group_by(grade) %>% 
  summarise(count_birds = n()) %>% 
  mutate(percent_bird = (count_birds/sum(count_birds))*100) %>% 
  select(grade, percent_bird) %>% 
  st_drop_geometry()

kable(bird_final, 
      digits = 2, 
      col.names = c("Grade", "Percent of bird observations"), 
      caption = "Percent of bird observations by HOLC redlining grade", 
      align = "l")
```

## Histogram of percent bird observations by HOLC grade

```{r}
#| code-fold: true
#| code-summary: "Reveal code"

ggplot(bird_final, aes(x = grade, y = percent_bird)) +
  geom_col(aes(fill = grade)) +
  scale_fill_manual(values=c("#91f366", "#3472db","#f0f061", "#f54242", "grey35")) +
  labs(title = "Percent of bird observations by HOLC grade",
       x = "HOLC Grade",
       y = "Bird observations (%)") +
  theme_minimal()
```
These findings do not match the results of Ellis-Soto et al. 2023, who find that historically redlined areas remain the most undersampled urban areas for bird biodiversity. Contrarily, the results of my histogram suggest that the highest percentage of bird observations occur in HOLC grade C and D. This result could be explained by the fact that we do not consider the total bird observations standardized by the size of the HOLC grade. That is, the area of the C and D grades might be larger than the A and B grades, thus the grades with larger areas will naturally have more bird observations. 

### Data Citations

```{r}
#| code-fold: true
#| code-summary: "Reveal code"

tribble(
  ~Data, ~Citation, ~Link,
  "Global Biodiversity Information Facility", "GBIF.org (13 October 2024) GBIF Occurence Download", "[Species API](https://techdocs.gbif.org/en/openapi/v1/species)",
  "HOLC Redlining Data", "Nelson, R. K., Winling, L, et al. (2023). Mapping Inequality: Redlining in New Deal America. Digital Scholarship Lab. https://dsl.richmond.edu/panorama/redlining.", "[Mapping Inequality](https://dsl.richmond.edu/panorama/redlining/data)",
  "EJScreen: Environmental Justice Screening and Mapping Tool", "United States Environmental Protection Agency. 2024 version. EJScreen. Retrieved: 10/4/24 from www.epa.gov/ejscreen", "[EJScreen Data](https://www.epa.gov/ejscreen/download-ejscreen-data)"
) %>% 
  kable()
```